---
title: "Assignment 5: Data Visualization"
author: "Prithveeraj Wavre"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
getwd()
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

NTL_LTER_processed <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

 Niwot_Ridge_litter <- 
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)
#2 
NTL_LTER_processed$sampledate <- ymd(NTL_LTER_processed$sampledate)
Niwot_Ridge_litter$collectDate <- ymd(Niwot_Ridge_litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
# Create a custom theme
mytheme <- 
  theme(
    plot.background = element_rect(fill = "lightgray", color = NA),  # Plot background color
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  # Plot title style
  )

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
# Plot total phosphorus by phosphate with separate aesthetics for lakes
plot <- ggplot(NTL_LTER_processed, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Total Phosphorus by Phosphate in Peter and Paul Lakes") +
  xlab("Phosphate (po4)") +
  ylab("Total Phosphorus (tp_ug)") +
  xlim(0, 50) +  # Adjust the limits for the x-axis to hide extreme values
  ylim(0, 50) +  # Adjust the limits for the y-axis to hide extreme values
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "top"
  ) + 
  mytheme

# Display the plot
print(plot)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

# Create the three separate boxplots
temp_plot <- ggplot(NTL_LTER_processed, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  ggtitle("Temperature by Month") +
  ylab("Temperature (°C)") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none"  # Remove legend for individual plot
  ) + 
  mytheme

tp_plot <- ggplot(NTL_LTER_processed, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  ggtitle("Total Phosphorus by Month") +
  ylab("Total Phosphorus") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none"  # Remove legend for individual plot
  ) + 
  mytheme

tn_plot <- ggplot(NTL_LTER_processed, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  ggtitle("Total Nitrogen by Month") +
  ylab("Total Nitrogen") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top"  # Show legend for this plot
  ) + 
  mytheme

# Combine the three plots into a single cowplot
combined_plot <- plot_grid(
  temp_plot, tp_plot, tn_plot,
  nrow = 3,
  align = "v",
  axis = "lr"
)

# Display the combined plot
print(combined_plot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The median for temperature falls in month of June for Paul lake and in the month of September for Peter lake. 
The median for Phosphorus and Nitrogen both, falls in the month of June for Paul and in the month of July for Peter lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
# Filter the dataset to include only the "Needles" functional group
needles_data <- Niwot_Ridge_litter %>% 
  filter(functionalGroup == "Needles")

# Create the plot
plot1 <- ggplot(needles_data, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  geom_line() +
  ggtitle("Dry Mass of Needle Litter by Date") +
  xlab("Collect Date") +
  ylab("Dry Mass") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "top"
  ) +
  mytheme

# Display the plot
print(plot1)

#7
# Create the plot with facets for NLCD classes
plot2 <- ggplot(needles_data, aes(x = collectDate, y = dryMass)) +
  geom_point() +
  geom_line() +
  ggtitle("Dry Mass of Needle Litter by Date") +
  xlab("Collect Date") +
  ylab("Dry Mass") +
  facet_wrap(~ nlcdClass, ncol = 1) +  # Facet by NLCD class with one column
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_text(color = "black"),
    legend.position = "none"  # Remove legend since facets are used
  ) +
  mytheme

# Display the plot
print(plot2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective as visually it is easier to identify the NLCD classes with facet wrap rather than just color difference. 